Worflow

### 2017-06-06 R/workflow.Rmd
workflow with se

```{r}
runZinb <- FALSE
runClus <- TRUE
NCORES <- 8
```

@@

```{r}
fn50 <- '../data/zinb_k50_batch.rda'
fn0 <- '../data/zinb_k0_batch.rda'
if (runZinb){
  mod = model.matrix( ~ batch)
  # K = 0
  print(system.time(zinb0 <- zinbFit(core, ncores = NCORES, K = 0, X = mod)))
  save(zinb0, file = f50)
  # K = 50
  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES, K = 50, X = mod)))
  save(zinb50, file = fn50)
}else{
  load(fn50)
  load(fn0)
}
```

@@

```{r rsec_50}
W = colData(se)[, grepl('^W', colnames(colData(se)))]
W = as.matrix(W)
fn = '../data/RSEC_W.rda'
if (runClus){
  print(system.time(ceObj <- RSEC(t(W), k0s = 4:15, 
                                  clusterFunction = "hierarchical01",
                                  minSizes = 5,
                                  ncores = NCORES,
                                  combineProportion = 0.7,
                                  mergeMethod = "locfdr", mergeCutoff = 0.01)))
  save(ceObj, file = fn)
}else{
  load(fn)
}
```

@@

```{r examineMergeClusters}
#re-does merging simpling to make plot 
mergeClusters(ceObj, mergeMethod = "locfdr", plotInfo="mergeMethod",cutoff = 0.01)
```

```{r heatmap }
plotHeatmap(ceObj, clusterSamplesData="dendrogramValue", breaks = .99)
```

```{r compareClusters}
plot(W, col = col_clus[as.character(clus.labels)], main = 'color = previous', pch = 20)
plot(W, col = as.factor(primaryCluster(ceObj)), main = 'color = new', pch = 20)
table(data.frame(previous = clus.labels, new = primaryCluster(ceObj)))
```

So, let's look at a heatmap on deviance residuals using heatmap.2 from gplots package with all the 1000 most variable genes.
```{r, eval=FALSE}
palDF = merged@clusterLegend[[1]]
pal = palDF[, 'color']
names(pal) = palDF[, 'name']

plotHeatmap(res, Colv = merged@dendro_samples,
          scale = 'none', trace = 'none', 
          col = colorRampPalette(c('blue', 'yellow'))(51),
          ColSideColors = pal[primaryClusterNamed(merged)],
          labCol = '', main = 'Deviance Residuals, 1000 most variable genes')
legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
       horiz = T, cex = 0.5)
```

## Get lineages
I used the unsupervised mode on W got from fitting zinbwave with batches in X and K = 4.

```{r slingshot}
fn = '../data/zinb_k4_batch.rda'
if (runZinb){
  cl_labs = primaryCluster(merged)
  batch_red = as.vector(batch)
  batch_red = batch_red[cl_labs != -1]
  cl_labs = cl_labs[cl_labs != -1]
  mod = model.matrix( ~ batch_red)
  zinb4 = zinbFit(core[,primaryCluster(merged) != -1],
                  X = mod, K = 4, ncores = NCORES)
  save(zinb4, file = fn)
}else{
  cl_labs = primaryCluster(merged)
  batch_red = as.vector(batch)
  batch_red = batch_red[cl_labs != -1]
  cl_labs = cl_labs[cl_labs != -1]
  mod = model.matrix( ~ batch_red)
  load(fn)
}
```

```{r}
X <- zinb4@W

lineages <- get_lineages(X, clus.labels = cl_labs, start.clus = 1)
curves <- get_curves(X, clus.labels = cl_labs, lineages = lineages)
plot_curves(X, cl_labs, curves, col.clus = unique(cl_labs))

print(lineages$lineage1)
print(lineages$lineage2)
print(lineages$lineage3)
print(lineages$lineage4)
```

@@

DE analysis using ZINB-Wave has not been implemented yet and we are still thinking about it. Would it make sense just to test the betas between clusters making the assumption that the genes are independent? Below is just a visualization of what we could have at the end of the worflow: a heatmap of the deviance residuals for DE genes between clusters.

```{r}
de = read.csv('../data/oe_markers.txt', stringsAsFactors = F, header = F)
de = de$V1

breaks = setBreaks(res[rownames(res) %in% de, ], 0.99)
heatmap.2(res[rownames(res) %in% de, ], breaks = breaks,
          Colv = merged@dendro_samples,
          scale = 'none', trace = 'none', 
          col = colorRampPalette(c('blue', 'yellow'))(51),
          ColSideColors = pal[primaryClusterNamed(merged)],
          labCol = '', main = 'Deviance Residuals, DE genes (Russell)')
legend("bottom", legend = names(pal), fill = pal, title = 'Clusters',
       horiz = T, cex = 0.5)
```

>>>>

```{r}
runZinb <- T
runClus <- T
NCORES <- 10
```

@@

```{r}
fn0 <- '../data/zinb_k0_batch.rda'
if (runZinb & !file.exists(fn0)){
  mod <- model.matrix( ~ batch)
  print(system.time(zinb0 <- zinbFit(core, ncores = NCORES,
                                     K = 0, X = mod)))
  save(zinb0, file = fn0)
}else{
  load(fn0)
}

fn50 <- '../data/zinb_k50_batch.rda'
if (runZinb & !file.exists(fn50)){
  mod <- model.matrix( ~ batch)
  print(system.time(zinb50 <- zinbFit(core, ncores = NCORES,
                                      K = 50, X = mod)))
  save(zinb50, file = fn50)
}else{
  load(fn50)
}
```

@@

```{r rsec_50}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
fn <- '../data/RSEC_W.rda'
if (runClus & !file.exists(fn)){
  print(system.time(ceObj <- RSEC(t(W), k0s = 4:15, 
                                  alphas = c(0.3), betas = c(0.9),
                                  clusterFunction = "hierarchical01",
                                  minSizes = 5,
                                  ncores = NCORES,
                                  combineProportion = 0.7,
                                  mergeMethod = "locfdr",
                                  mergeCutoff = 0.01)))
  save(ceObj, file = fn)
}else{
  load(fn)
}
```

@@

FP: Elizabeth, we are working with the W here, does the locfdr make sense in this context?

```{r examineMergeClusters}
#re-does merging simpling to make plot 
mergeClusters(ceObj, mergeMethod = "locfdr",
              plotInfo = "mergeMethod", cutoff = 0.05)
```

```{r heatmap }
plotHeatmap(ceObj, clusterSamplesData = "dendrogramValue", 
            breaks = .99)
```

```{r compareClusters}
plot(fit$points, col = col_clus[as.character(clus.labels)],
     main = 'MDS W, color = published clusters', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
```

```{r}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, 'color']
names(pal) <- palDF[, 'name']
plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
     main = 'MDS W, color = our clusters', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
```

```{r}
table(data.frame(published = clus.labels,
                 ours = primaryCluster(ceObj)))
```

So, let's look at a heatmap on normalized values.

```{r}
sampleData <- data.frame(ours = primaryCluster(ceObj))

plotHeatmap(assays(se)$normalizedValues,
            main = 'Normalized values, 1000 most variable genes',
            clusterSamplesData = ceObj@dendro_samples,
            sampleData = as.matrix(sampleData))
```

## Get lineages

The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We run slingshot in the supervised and unsupervised mode and we try k=2, k=3, k=4 in W.

FP: We need to work on the cluster coloring (!) and the matching with the clusters you found for the paper with Russell, maybe by changing the parameters in clusterMany?

```{r}
Kvec <- c(2, 3, 4)
```

### Use previous W, no re-fitting 
#### Unsupervised

```{r slingshot_unsup}
our_cl <- as.factor(primaryClusterNamed(ceObj))
cl = our_cl[our_cl != "-1"]
for (k in Kvec){
  X <- W[our_cl != "-1", 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
}
```

#### Supervised

```{r slingshot_sup}
for (k in Kvec){
  X <- W[our_cl != "-1", 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1",
                           end.clus = c("m9", "m6"))
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
}
```

### Re-fitting zinbwave
#### Unsupervised

```{r refit_zinb}
mod <- model.matrix(~ batch[our_cl != "-1"])
fn <- '../data/refit_zinbwave_slingshot.rda'
if (run & !file.exists(fn)){
  zinbList <- lapply(Kvec, function(k){
    zinbFit(core[, our_cl != "-1"], X = mod, K = k, ncores = NCORES)
  })
  save(zinbList, file = fn)
}else{
  load(fn)
}
```

```{r slingshot_unsup_refit}
for(k in Kvec) {
  X <- getW(zinbList[[k - 1]])[, 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "m1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
}
```

#### Supervised

```{r slingshot_sup_refit}
for(k in Kvec){
  X <- getW(zinbList[[k - 1]])[, 1:k]

  lineages <- get_lineages(X, clus.labels = our_cl, start.clus = "m1",
                           end.clus = c("m9", "m6"))
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
}
```

@@

Here is the kind of plots we want to be able to have

```{r}
de <- read.csv('../data/oe_markers.txt', 
              stringsAsFactors = F, header = F)
de <- de$V1
nv <- assays(se)$normalizedValues
plotHeatmap(nv[rownames(nv) %in% de, ],
            main = 'Normalized values, 1000 most variable genes',
            clusterSamplesData = ceObj@dendro_samples,
            sampleData = as.matrix(sampleData$ours))
```
